---
title: "R Basics"
author: "Paulo Cirino Ribeiro Neto"
date: "26/01/2017"
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, prompt = TRUE, results = TRUE, highlight = TRUE)
```

## Course Outline {.build .emphasized}
- Syntax, data structures and operators;
- Useful R base functions;
- Conditional and control flow statements;
- Creating custom Functions;
- The **apply** functions;

# Syntax, data structures and operators;

## Arithmetics with R {.build .emphasized}
### Addition Operator: '+'
```{r, eval = TRUE}
5 + 2
```

###	Subtraction Operator: '-' 
```{r, eval = TRUE}
5 - 2
```

## Arithmetics with R {.build .emphasized}
### Multiplication Operator: '*'
```{r, eval = TRUE}
5 * 2
```

### Division Operator: '/'
```{r, eval = TRUE}
5 / 2
```

## Arithmetics with R {.build .emphasized}
### Exponentiation Operator: '**' or '^'
```{r, eval = TRUE}
5 ** 2
5 ^ 2
```

## Arithmetics with R {.build .emphasized}
### Modulus Operator: '%%'
```{r, eval = TRUE}
5 %% 2
```

### Integer Division Operator: '%/%' 
```{r, eval = TRUE}
5 %/% 2
```

## Data Types {.build .emphasized}
### Integer:
  - Numbers belonging to ${\rm I\!N}$ Set;
  
### Numeric:
  - Numbers belonging to ${\rm I\!R}$ Set;
  
### Complex:
  - Numbers belonging to ${\rm I\!C}$ Set;
  
## Data Types {.build .emphasized}
### Character:
  - Characters or Strings
  
### Factor:
  - Categorical Variables
  
### Dates:
  - Formated Dates

## Data Structures {.build}
### Vectors :
  - It is a sequence of any one of the *data types*, it also can grow dynamically.

```{r, echo = TRUE, eval = TRUE}
c('foo', 'bar', 'baz')
```

```{r, echo = TRUE, eval = TRUE}
c(-3.2, 4.6, 5.1)
```

```{r, echo = TRUE, eval = TRUE}
1:10
```
  
## Data Structures {.build}
### Matrices : 
  - A 2 dimension array of any one of the data types that has got pre-established size.
```{r, echo = TRUE, eval = TRUE}
matrix(data = 1:6, ncol = 3, nrow = 2)
```

```{r, echo = TRUE, eval = TRUE}
matrix(data = c('foo', 'baz', 'bar', 'faz'),
       ncol = 2, nrow = 2)
```
  
## Data Structures {.build}
### Arrays : 
  - An multidimensional array of any one of the data types that has got pre-established size.
```{r}
array(data = 1:12, dim = c(2,3,2))
```
  
## Data Structures {.build}
### Data Frames :
  - A 2 dimension array that can hold multiple data types, each one in a different column.
```{r}
data.frame(Name = c('John', 'Jack', 'Jose'),
           Age = c(20, 25, 22),
           Height = c(1.78, 1.92, 1.71))
```
  
## Data Structures {.build}
### Lists : 
  - A collction of many different data types that can grow dynamically.
```{r}
list(Supermarket = data.frame(
  Product = c('Soup', 'Potatoes', 'Tooth Paste'),
  Quantity = c(3, 20, 2)),
  Date = Sys.Date())
```

## Variable Assignment {.build}
### Left Assignment : '<-'
```{r, eval = TRUE}
numbers <- c(3.2, 5.7, -12.1)
numbers
```

```{r, eval = TRUE}
name <- 'Jack'
name
```

## Variable Assignment {.build}
### Left Assignment : '<<-'
```{r, eval = TRUE}
number <<- 13
number
```

```{r, eval = TRUE}
cols <<- factor(c('blue', 'black', 'blue'))
cols
```

## Variable Assignment {.build}
### Difference of '<<-' and '<-' :

The '<<-' will check parent environments for an existing variable, if a variable is not found then its value is redefined, otherwise it take the global environment value.

```{r, eval = TRUE, error = TRUE, warning = TRUE}
try(c <<- factor(c('blue', 'black', 'blue')))
c
```

```{r, eval = TRUE}
try(c <- factor(c('blue', 'black', 'blue')))
c
```

## Variable Assignment {.build}
### Difference of '<<-' and '<-' :

```{r, eval = TRUE, error = TRUE, warning = TRUE}
pi
```

```{r, eval = TRUE}
try(pi <<- -19)
pi
```

```{r, eval = TRUE}
try(pi <- -19)
pi
```

## Variable Assignment {.build}
### Right Assignment : '->' and '->>'
```{r, eval = TRUE}
letters
```

```{r, eval = TRUE}
try(1234 ->> letters)
letters
```

```{r, eval = TRUE}
try(1234 -> letters)
letters
```

## Variable Assignment {.build}
### Right Assignment : '->' and '->>'
```{r, eval = TRUE}
c(TRUE, TRUE, FALSE) ->> Bools
Bools
```

```{r, eval = TRUE}
c(FALSE, FALSE, TRUE) ->> Bools
Bools
```

## Variable Assignment {.build}
### Equal Assignment : '='
```{r, eval = TRUE}
x <- 10
print(args(sample))
```

```{r, eval = TRUE}
sample(x = 6)
```

```{r, eval = TRUE}
x
```

## Variable Assignment {.build}
### Equal Assignment : '='
```{r, eval = TRUE}
sample(x <- 3 )
```

```{r, eval = TRUE}
x
```

## Variable Assignment {.build}
  - The '->', '<-' and '=' are assign into the environment in which they are evaluated (inside a funciton or a script).
  - The '->>' and '<<-' ared assigned globally.
  - The '<-' is the correct assignment operator for variables.
```{r}
a <- 15
b <- 'character string'
```


## Variable Assignment {.build}
  - The '=' is the correct assignment operator for subexpressions in a braced list of expressions.
```{r}
listVar <- list(var1 = 12, var2 = 'abc')
print(x = listVar)
``` 


## Relational Operators
- Less than : '<'
- Less than or Equal to : '<='
- Greater than : '>'
- Greater than or Equal to : '>='
-	Exactly Equal to : '=='
-	Not Equal to : '!='

## Relational Operators
```{r}
TRUE == 1
```

```{r}
12 != 'b'
```

```{r}
12 >= 12
```

## Relational Operators on Vectors
```{r}
10 > 9:11
```

```{r}
rep(10, 3) == 9:11
```

```{r}
9:11 == rep(10, 3)
```

## Logical Operators
-	Not : '!'
```{r}
!TRUE
```

```{r}
!2
```

```{r}
!FALSE
```

## Logical Operators
### And : '&' or '&&'
```{r}
TRUE & 1
``` 

```{r}
( (2:4) >= 3) & ( (2:4) <= 3)
```

```{r}
( (2:4) >= 3) && ( (2:4) <= 3)
```


## Logical Operators
### Or : '|' or '||'
```{r}
( (2:4) >= 3) | ( (2:4) == 2)
```

```{r}
( (2:4) >= 3) || ( (2:4) == 2)
```

```{r}
( (2:4) >= 3) | ( (2:4) == 3)
```

```{r}
( (2:4) >= 3) || ( (2:4) == 3)
```

## Selection Operators
### Select : '[]'
```{r}
LETTERS
```

```{r}
LETTERS[5]
```

```{r}
LETTERS[5:10]
```

## Selection Operators
### Select : '[]'
```{r}
LETTERS[-1]
```

```{r}
LETTERS[-1:-10]
```

## Selection Operators
### Select : '[]'
```{r}
auxVar <- c('apple', 'orange', 'ávocado')
auxVar
```

```{r}
auxVar[1] <- 'strawberry'
auxVar
```

```{r}
auxVar[3] <- 'grapes'
auxVar
```

## Selection Operators
### Select : '[]'
```{r}
auxVar[4] <- 'blueberry'
auxVar
```

```{r}
auxVar[7] <- 'water melon'
auxVar
```

## Selection Operators
### Select : '$'

```{r}
auxList <- list( names = c('Paulo', 'Cirino'),
                 numbers = c(12, 14, 15))
auxList$names
```

```{r}
auxList$numbers
```

## Miscellaneous Operators
- Range : ':'
```{r}
1:10
```

- In : '%in%'
```{r}
5 %in% 1:10
```

```{r}
40 %in% 1:10
```

- Help : '?'
```{r}
?class
```

## Pro Tips
- Always use the __?__ operator and the **help()** function when dealing with unknown functions or packages;
- Always use **<-** operator when assignning a variable in the program a value, exe : **a <- 10** ;
- Always use the **=** operator when assignning a variable in a function call, exe : **is.na(x = a)**;
- Avoid using **->** operator, because it makes the code messy;
- Always use the **\$** operator to select named columns of data frame, exe : DF$X1 ;
- Always use the **[]** operator to select row or columns when using operation, exe : **DF[DF$X1 == 1, ]**;

# Useful R base functions

## Important Functions
### class : Checks the class of a variable.
```{r}
class(1234)
```

### help : Opens package or funciton documentation.
```{r, eval = FALSE}
help(lm)
```

### summary : Returns Variable Summary.
```{r}
summary(1:10)
```


## Important Functions
### Sys.time : Returns a Time Stamp.
```{r}
Sys.time()
```

### sample : Generates Sample integers
```{r, eval = FALSE}
sample(10, 3)
```

### rnorm : Generates Normal Distribution.
```{r}
rnorm(10)
```

## Important Functions
### print : Prints String as is.
```{r}
print('R is super \'cools\'')
```

### cat : Prints C like function expressions.
```{r, eval = FALSE}
cat('R is super \'cools\'')
```

### try : Try a function but keeps going
```{r}
try(asdfasd * 1234213 / 0)
```

## Important Functions
### library : Loads a package.
```{r, eval = FALSE}
library('Rcpp')
```

### require : Try loading a package.
```{r, eval = FALSE}
require('DT')
```

### install.packages : Install a package
```{r, eval = FALSE}
install.packages('ggplot2')
```

### View : View a Data Frame or Matrix
```{r, eval = FALSE}
View(iris)
```

## Important Functions
### source : Source R file
```{r, eval = FALSE}
source('Code.R')
```

### paste : Paste things as characters
```{r, eval = FALSE}
paste('R', ' is like a ' 10)
```

### order : Returns order of a vector
```{r, eval = FALSE}
order(c(10, -3, 1)) # returns [2 3 1]
```

### order : Returns order of a vector
```{r, eval = FALSE}
order(c(10, -3, 1)) # returns [2 3 1]
```

### order : Returns order of a vector
```{r, eval = FALSE}
rm( list = c('Data1', 'Data2') )
```


## Math Functions

### t : Transpose
```{r}
t(matrix(1:9, ncol = 3, nrow = 3))
```

### mean : Calculates mean
```{r}
mean(1:3)
```

### max : Calculates max
```{r}
max(1:3)
```

## Math Functions

### sum : Sums a vector
```{r}
sum (1:3)
```

### round : Rounds Closest
```{r}
round (3.14, 1)
```

### ceil : Rounds up.
```{r}
ceiling (3.14)
```

### floor : Rounds down
```{r}
floor (3.14)
```


## Data Frames
### nrow : Counts Number of Rows
```{r}
nrow(iris)
```

### ncol : Counts Number of Columns
```{r}
ncol(iris)
```

### dim : Returns Number of Dimensions
```{r}
dim(iris)
```

## Data Frames
### tail : Last Data Frame Observations
```{r}
tail(iris, 3)
```

### head : Fist Data Frame Observations
```{r}
head(iris)
```

### colnames : Column names of Data Frames or Matrixes
```{r}
colnames(iris)
```

## List
```{r}
myList <- list(a = 1:10, b = LETTERS[10:15])
```

### length : Length of Object
```{r}
length(myList)
```

### names : Objects named inner variables
```{r}
names(myList)
```


## Pro Tips
- Always use **dim()**, **class()**, **names()**, **head()**, **summary()** beforehand when dealing with unknown dataset;

- When creating scripts try using 

# Conditional and control flow statements;
# Creating custom Functions;
# The **apply** functions;
